---
title: 'predicting mRSI (fatigue) with Braking RFD and more?'
author: "Jordyn Maurer"
date: "2/11/2025"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


## Introduction
Fencing, BB, WBB, MBB, MLAX, WLAX, Golf (December/January), FB full season CMJ metrics. 

## Data Cleaning 

```{r}
# updating file path
setwd("C:/Users/jordy/OneDrive/Documents/Coach_scott/mRSI_BRFD")

# loading in the library
library(dplyr)
```

```{r}
BB = read.csv("BB.csv")
Fencing = read.csv("Fencing.csv")
Golf = read.csv("Golf.csv")
MBB = read.csv("MLAX.csv")
WBB = read.csv("WBB.csv")
WLAX = read.csv("WLAX.csv")
MLAX = read.csv("MLAX.csv")
FB = read.csv("FB_CMJ_Data.csv")
```

```{r}
# selecting only necessary columns and creating a team column for baseball
BB_clean <- BB %>%
  select(TestId, Segment, Date, Braking.RFD, mRSI, Braking.Net.Impulse, Peak.Braking.Force, Relative.Propulsive.Impulse, Peak.Braking.Velocity, Peak.Propulsive.Force, Peak.Propulsive.Power, Peak.Braking.Force, Avg..Braking.Power) %>% 
  mutate(Team = "BB")

# selecting only necessary columns and creating a team column for fencing
Fencing_clean <- Fencing %>%
  select(TestId, Segment, Date, Braking.RFD, mRSI, Braking.Net.Impulse, Peak.Braking.Force, Relative.Propulsive.Impulse, Peak.Braking.Velocity, Peak.Propulsive.Force, Peak.Propulsive.Power, Peak.Braking.Force, Avg..Braking.Power) %>% 
  mutate(Team = "Fencing")

# selecting only necessary columns and creating a team column for golf
Golf_clean <- Golf %>%
  select(TestId, Segment, Date, Braking.RFD, mRSI, Braking.Net.Impulse, Peak.Braking.Force, Relative.Propulsive.Impulse, Peak.Braking.Velocity, Peak.Propulsive.Force, Peak.Propulsive.Power, Peak.Braking.Force, Avg..Braking.Power) %>% 
  mutate(Team = "Golf")

# selecting only necessary columns and creating a team column for men's basketball
MBB_clean <- MBB %>%
  select(TestId, Segment, Date, Braking.RFD, mRSI, Braking.Net.Impulse, Peak.Braking.Force, Relative.Propulsive.Impulse, Peak.Braking.Velocity, Peak.Propulsive.Force, Peak.Propulsive.Power, Peak.Braking.Force, Avg..Braking.Power) %>% 
  mutate(Team = "MBB")

# selecting only necessary columns and creating a team column for women's basketball
WBB_clean <- WBB %>%
  select(TestId, Segment, Date, Braking.RFD, mRSI, Braking.Net.Impulse, Peak.Braking.Force, Relative.Propulsive.Impulse, Peak.Braking.Velocity, Peak.Propulsive.Force, Peak.Propulsive.Power, Peak.Braking.Force, Avg..Braking.Power) %>% 
  mutate(Team = "WBB")

# selecting only necessary columns and creating a team column for WLAX
WLAX_clean <- WLAX %>%
  select(TestId, Segment, Date, Braking.RFD, mRSI, Braking.Net.Impulse, Peak.Braking.Force, Relative.Propulsive.Impulse, Peak.Braking.Velocity, Peak.Propulsive.Force, Peak.Propulsive.Power, Peak.Braking.Force, Avg..Braking.Power) %>% 
  mutate(Team = "WLAX")

# selecting only necessary columns and creating a team column for MLAX
MLAX_clean <- MLAX %>%
  select(TestId, Segment, Date, Braking.RFD, mRSI, Braking.Net.Impulse, Peak.Braking.Force, Relative.Propulsive.Impulse, Peak.Braking.Velocity, Peak.Propulsive.Force, Peak.Propulsive.Power, Peak.Braking.Force, Avg..Braking.Power) %>% 
  mutate(Team = "MLAX")

# selecting only necessary columns and creating a team column for FB
FB_clean <- FB %>%
  select(TestId, Segment, Date, Braking.RFD, mRSI, Braking.Net.Impulse, Peak.Braking.Force, Relative.Propulsive.Impulse, Peak.Braking.Velocity, Peak.Propulsive.Force, Peak.Propulsive.Power, Peak.Braking.Force, Avg..Braking.Power) %>% 
  mutate(Team = "FB")
```


```{r}
# combining all CSVs into one dataset
all_teams_data <- bind_rows(BB_clean, Fencing_clean, Golf_clean, MBB_clean, 
                             WBB_clean, WLAX_clean, MLAX_clean, FB_clean)
```

```{r}
# previewing the dataset
head(all_teams_data)
summary(all_teams_data)
```

## Exploratory Analysis of mRSI and Braking RFD

```{r}
# loading in the library to create visualizations
library(ggplot2)

# creating an initial scatter plot of Braking RFD and mRSI
ggplot(all_teams_data, aes(x = Braking.RFD, y = mRSI)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = 'blue') +
  facet_wrap(~ Team)
```

From the initial scatter plots, there may be some linear relationship between Braking RFD and mRSI, but there seems to be a lot of variaton. 

```{r}
# checking the correlation coefficient for all teams
all_teams_data %>%
  summarise(correlation = cor(Braking.RFD, mRSI, use = "complete.obs"))

all_teams_data %>%
  group_by(Team) %>%
  summarise(correlation = cor(Braking.RFD, mRSI, use = "complete.obs"))
```

For all teams, the correlation coefficient is about .47, meaning there may not necessarily be any relationship between mRSI and BRFD in the larger data set. After splitting the observations by team, Golf seems to have the highest correlation, with .71 as its coefficient, but it also has the least amount of observations. Football, with the most observations recorded in the data, shows the smallest correlation between mRSI and BRFD. I'm going to create an initial linear regression to see if anything is significant.`

```{r}
# linear model with all team data
lm_braking_rfd <- lm(mRSI ~ Braking.RFD, data = all_teams_data)

summary(lm_braking_rfd)
```

This initial model shows BRFD may be a strong predictor of mRSI. However, the R-Squared value shows that only about 22% of the variation in mRSI can be explained by BRFD. This suggests that while a relationship between the two variables may exist, it's definitely not the only factor influencing mRSI. To see if there are similar results, I'm going to run the regression across teams. 

```{r}
# BB
BB_rfd <- lm(mRSI ~ Braking.RFD, data = BB_clean)
summary(BB_rfd)

# Golf
Golf_rfd <- lm(mRSI ~ Braking.RFD, data = Golf_clean)
summary(Golf_rfd)

# FB
FB_rfd <- lm(mRSI ~ Braking.RFD, data = FB_clean)
summary(FB_rfd)
```

Splitting this apart into separate linear models confirmed the findings from the correlation findings. The model for Golf explained the most variation in mRSI while the model for Football showed the model explaining only 13% variation, an extremely low value. There is definitely a relationship between Braking RFD and mRSI, but Braking RFD isn't the only predictor for mRSI. 

## Analyzing mRSI as a model of other metrics as well as Braking RFD

Before I create the model I want to check the behavior of the KPIs to see if some of them should be transformed. 

```{r}
library(ggplot2)

# Histogram for Braking RFD
ggplot(all_teams_data, aes(x = Braking.RFD)) +
  geom_histogram(binwidth = 1000, fill = "maroon", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Braking RFD", x = "Braking RFD", y = "Frequency") +
  theme_minimal()

# Histogram for mRSI
ggplot(all_teams_data, aes(x = mRSI)) +
  geom_histogram(binwidth = 0.05, fill = "maroon", color = "black", alpha = 0.7) +
  labs(title = "Histogram of mRSI", x = "mRSI", y = "Frequency") +
  theme_minimal()

# Histogram for Peak Braking Force
ggplot(all_teams_data, aes(x = Peak.Braking.Force)) +
  geom_histogram(binwidth = 50, fill = "maroon", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Peak Braking Force", x = "Peak Braking Force", y = "Frequency") +
  theme_minimal()

# Histogram for Relative Propulsive Impulse
ggplot(all_teams_data, aes(x = Relative.Propulsive.Impulse)) +
  geom_histogram(binwidth = 0.1, fill = "maroon", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Relative Propulsive Impulse", x = "Relative Propulsive Impulse", y = "Frequency") +
  theme_minimal()

# Histogram for Peak Braking Velocity
ggplot(all_teams_data, aes(x = Peak.Braking.Velocity)) +
  geom_histogram(binwidth = 0.1, fill = "maroon", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Peak Braking Velocity", x = "Peak Braking Velocity", y = "Frequency") +
  theme_minimal()

# Histogram for Peak Propulsive Force
ggplot(all_teams_data, aes(x = Peak.Propulsive.Force)) +
  geom_histogram(binwidth = 50, fill = "maroon", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Peak Propulsive Force", x = "Peak Propulsive Force", y = "Frequency") +
  theme_minimal()

# Histogram for Avg. Braking Power
ggplot(all_teams_data, aes(x = Avg..Braking.Power)) +
  geom_histogram(binwidth = 50, fill = "maroon", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Average Braking Power", x = "Average Braking Power", y = "Frequency") +
  theme_minimal()

```


```{r}
# finding the best model for mRSI
numeric_data <- all_teams_data %>%
  select_if(is.numeric)
head(numeric_data)

library(leaps) 

# forward stepwise selection
regfit.fwd <- regsubsets(mRSI ~ ., data = numeric_data, nvmax = 9, method = "forward")

# backward stepwise selection
regfit.bwd <- regsubsets(mRSI ~ ., data = numeric_data, nvmax = 9, method = "backward")
```

```{r}
# Forward Selection Summary
fwd.summary <- summary(regfit.fwd)

# RSS
plot(fwd.summary$rss, xlab = "Number of Variables", ylab = "RSS", type = "l", main = "Optimal Number of Variables Based on RSS with Forward Selection", cex.main = 0.75)
which.min(fwd.summary$rss)
points(which.min(fwd.summary$rss), min(fwd.summary$rss), col = "red", cex = 2, pch = 20)

# adjusted Rsq
plot(fwd.summary$adjr2, xlab = "Number of Variables", ylab = "Adjusted RSq", type = "l", main = "Optimal Number of Variables Based on Adjusted R-Squared with Forward Selection", cex.main = 0.75)
which.max(fwd.summary$adjr2)
points(which.max(fwd.summary$adjr2), max(fwd.summary$adjr2), col = "red", cex = 2, pch = 20)

# BIC 
plot(fwd.summary$bic, xlab = "Number of Variables", ylab = "BIC", type = "l", main = "Optimal Number of Variables Based on BIC with Forward Selection", cex.main = 0.8)
which.min(fwd.summary$bic)
points(which.min(fwd.summary$bic), min(fwd.summary$bic), col = "red", cex = 2, pch = 20)

# Mallows' Cp
plot(fwd.summary$cp, xlab = "Number of Variables", ylab = "Mallows' Cp", type = "l", main = "Optimal Number of Variables Based on Mallows' Cp with Forward Selection", cex.main = 0.8)
which.min(fwd.summary$cp)
points(which.min(fwd.summary$cp), min(fwd.summary$cp), col = "red", cex = 2, pch = 20)
```

```{r}
# running a model with all variables
lm_all <- lm(mRSI ~ ., data = numeric_data)
summary(lm_all)
```

```{r}
# checking for multicollinearity
library(car)
vif(lm_all)
```

As expected, a lot of these variables are correlated. So, even though they may be strong predictors of mRSI, their multicollinearity may inflate the model. So, we'll have to continue with different methods to best predict mRSI.


## Checking Predictive Power of BRFD and Injuries (Football)
```{r}
# reading in the injury reports
aug_inj = read.csv("Injury Report August.csv")
sep_inj = read.csv("Injury Report September.csv")
oct_inj = read.csv("October Injury Report.csv")
nov_inj = read.csv("November Inj.csv")

# filtering and cleaning each dataset
## august
aug_inj_fb <- aug_inj %>%
  filter(Organizations == "Football") %>%
  select(-Teams, -ProblemReportedBy, -IncidentReportCreatedBy)

head(aug_inj_fb)

## september
sep_inj_fb <- sep_inj %>%
  filter(Organizations == "Football") %>%
  select(-Teams, -ProblemReportedBy, -IncidentReportCreatedBy) %>%
  rename(ID = X)

head(sep_inj_fb)

## october 
oct_inj_fb <- oct_inj %>%
  filter(Organizations == "Football") %>%
  select(-Teams, -ProblemReportedBy, -IncidentReportCreatedBy)

head(oct_inj_fb)

## november 
nov_inj_fb <- nov_inj %>%
  filter(Organizations == "Football") %>%
  select(-ProblemReportedBy, -IncidentReportCreatedBy) %>%
  rename(ID = Teams)

head(nov_inj_fb)
```

```{r}
# joining the four months of injuries into one data frame 
combined_fb_inj <- bind_rows(aug_inj_fb, sep_inj_fb, oct_inj_fb, nov_inj_fb) 
head(combined_fb_inj)

# seeing the unique IDs
unique_ID_count <- combined_fb_inj %>%
  summarise(unique_count = n_distinct(ID))
print(unique_ID_count)

unique_ID <- combined_fb_inj %>%
  distinct(ID)
print(unique_ID)
```

```{r}
# joining the force plate data and the injury data
merged_fb <- combined_fb_inj %>%
  left_join(FB_clean, by = c("ID" = "Segment"))

head(merged_fb)
```

```{r}
# identifying the types of injuries to filter
types_of_injuries <- merged_fb %>%
  distinct(InjuryType)
print(types_of_injuries)
```

```{r}
# removing concussions and fractures from the data 
merged_fb <- merged_fb %>%
  filter(InjuryType != "Concussion") %>%
  filter(InjuryType != "Fracture") %>%
  filter(InjuryType != "Glenoid Labrum Tear (SLAP)")

# counting the new amount of unique IDs
new_unique_ID_count <- merged_fb %>%
  summarise(unique_count = n_distinct(ID))
print(new_unique_ID_count)

new_unique_ID <- merged_fb %>%
  distinct(ID)
print(new_unique_ID)
```

```{r}
# visualizing BRFD over time for these injured FB players
library(ggplot2)

str(merged_fb$Date)
merged_fb$Date <- as.Date(merged_fb$Date, format = "%m/%d/%Y")
```

```{r}
# Filter data for each injury type
sprain_data <- merged_fb %>% filter(InjuryType == "Sprain")
strain_data <- merged_fb %>% filter(InjuryType == "Strain")
meniscus_data <- merged_fb %>% filter(InjuryType == "Meniscus Tear")

# Plot each injury type separately
# Sprain
ggplot(sprain_data, aes(x = Date, y = Braking.RFD, color = as.factor(ID), group = ID)) +
  geom_line() +
  labs(title = "Braking RFD Over Time for Sprain", x = "Date", y = "Braking RFD", color = "ID") +
  theme_minimal() +
  scale_x_date(date_labels = "%Y-%m-%d")

# Strain
ggplot(strain_data, aes(x = Date, y = Braking.RFD, color = as.factor(ID), group = ID)) +
  geom_line() +
  labs(title = "Braking RFD Over Time for Strain", x = "Date", y = "Braking RFD", color = "ID") +
  theme_minimal() +
  scale_x_date(date_labels = "%Y-%m-%d")

# Meniscus Tear
ggplot(meniscus_data, aes(x = Date, y = Braking.RFD, color = as.factor(ID), group = ID)) +
  geom_line() +
  labs(title = "Braking RFD Over Time for Meniscus Tear", x = "Date", y = "Braking RFD", color = "ID") +
  theme_minimal() +
  scale_x_date(date_labels = "%Y-%m-%d")
```

## Adding a Marker for Problem Date

```{r}
# changing the format of "ProblemDate"
library(lubridate)
sprain_data$ProblemDate <- mdy(substr(sprain_data$ProblemDate, 5, 15))
strain_data$ProblemDate <- mdy(substr(strain_data$ProblemDate, 5, 15))
meniscus_data$ProblemDate <- mdy(substr(meniscus_data$ProblemDate, 5, 15))
```

```{r}
unique_ids <- unique(sprain_data$ID)

set.seed(123)
group1_ids <- sample(unique_ids, size = length(unique_ids) / 2)
group2_ids <- setdiff(unique_ids, group1_ids)

sprain_group1 <- sprain_data %>% filter(ID %in% group1_ids)
sprain_group2 <- sprain_data %>% filter(ID %in% group2_ids)
```


```{r}
max_problem_date <- max(as.Date(sprain_data$ProblemDate))

ggplot(sprain_group1, aes(x = Date, y = Braking.RFD, color = as.factor(ID), group = ID)) +
  geom_line() +
  geom_vline(aes(xintercept = as.Date(ProblemDate), color = as.factor(ID)), linetype = "dashed") +
  labs(title = "Braking RFD Over Time for Sprain (Group 1)", x = "Date", y = "Braking RFD", color = "ID") +
  theme_minimal() +
  scale_x_date(date_labels = "%Y-%m-%d", limits = c(min(sprain_data$Date), max_problem_date)) +
   theme(plot.title = element_text(size = 10))


ggplot(sprain_group2, aes(x = Date, y = Braking.RFD, color = as.factor(ID), group = ID)) +
  geom_line() +
  geom_vline(aes(xintercept = as.Date(ProblemDate), color = as.factor(ID)), linetype = "dashed") +
  labs(title = "Braking RFD Over Time for Sprain (Group 2)", x = "Date", y = "Braking RFD", color = "ID") +
  theme_minimal() +
  scale_x_date(date_labels = "%Y-%m-%d", limits = c(min(sprain_data$Date), max_problem_date)) +
  theme(plot.title = element_text(size = 10))

```

```{r}
max_problem_date_strain <- max(as.Date(strain_data$ProblemDate))

ggplot(strain_data, aes(x = Date, y = Braking.RFD, color = as.factor(ID), group = ID)) +
  geom_line() +
  geom_vline(aes(xintercept = as.Date(ProblemDate), color = as.factor(ID)), linetype = "dashed") +
  labs(title = "Braking RFD Over Time for Strain)", x = "Date", y = "Braking RFD", color = "ID") +
  theme_minimal() +
  scale_x_date(date_labels = "%Y-%m-%d", limits = c(min(strain_data$Date), max_problem_date)) +
  theme(plot.title = element_text(size = 10))


max_problem_date_meniscus <- max(as.Date(meniscus_data$ProblemDate))

ggplot(meniscus_data, aes(x = Date, y = Braking.RFD, color = as.factor(ID), group = ID)) +
  geom_line() +
  geom_vline(aes(xintercept = as.Date(ProblemDate), color = as.factor(ID)), linetype = "dashed") +
  labs(title = "Braking RFD Over Time for Meniscus Tear)", x = "Date", y = "Braking RFD", color = "ID") +
  theme_minimal() +
  scale_x_date(date_labels = "%Y-%m-%d", limits = c(min(meniscus_data$Date), max_problem_date)) +
  theme(plot.title = element_text(size = 10))
```

# For athletes with data before and after a reported injury, is there a statistically difference in BRFD?

```{r}
# FB134, FB107, FB21 before and after injury
merged_fb$ProblemDate <- mdy(substr(merged_fb$ProblemDate, 5, 15))

injury_data <- merged_fb %>%
  filter(ID %in% c("FB134", "FB107", "FB21")) %>%
  mutate(
    InjuryStatus = ifelse(Date < ProblemDate, "Pre_Injury", "Post_Injury")
  )

head(injury_data)
```

```{r}
library(tidyr)
# calculate the average Braking.RFD for each injury status (Pre and Post) per athlete
injury_data_diff <- injury_data %>%
  group_by(ID, InjuryStatus) %>%
  summarise(average_brfd = mean(Braking.RFD, na.rm = TRUE)) %>%
  spread(key = InjuryStatus, value = average_brfd) %>%
  mutate(difference = `Post_Injury` - `Pre_Injury`)

# preview the calculated differences
head(injury_data_diff)

# checking normality
shapiro_test_result <- shapiro.test(injury_data_diff$difference)
shapiro_test_result
```

Normality assumption passes. 

```{r}
# perform the paired t-test
t_test_result <- t.test(injury_data_diff$difference, mu = 0)
t_test_result
```

Shows no statistically significant difference in pre- and post- injury BRFD in these three athletes. Although FB21 had a pre-injury average BRFD of 8001.012 and post-injury of 4675.390. 