---
title: "Strength_Work_Feb_13.Rmd"
output:
  html_document: default
  word_document: default
date: "2025-02-13"
---

```{r}
knitr::opts_chunk$set(echo = FALSE)
```


```{r}
# Load necessary library
library(dplyr)
library(readr)

# Load dataset (update path as needed)
df <- read_csv("MLAX_Jan.csv")

# Inspect column names to find braking velocity column
colnames(df)

# Define the peak braking velocity column
peak_braking_velocity_col <- "Peak Braking Velocity"
```
```{r}
# Load necessary libraries
library(dplyr)
library(readr)

# Load dataset (update the path as needed)
df <- read_csv("MLAX_Jan.csv")

# Define the peak braking velocity column
peak_braking_velocity_col <- "Peak Braking Velocity"

# Summary statistics BEFORE exclusion
summary_before <- df %>%
  summarise(
    Total_Entries = n(),
    Mean_Braking_Velocity = mean(get(peak_braking_velocity_col), na.rm = TRUE),
    Median_Braking_Velocity = median(get(peak_braking_velocity_col), na.rm = TRUE),
    Min_Braking_Velocity = min(get(peak_braking_velocity_col), na.rm = TRUE),
    Max_Braking_Velocity = max(get(peak_braking_velocity_col), na.rm = TRUE),
    SD_Braking_Velocity = sd(get(peak_braking_velocity_col), na.rm = TRUE)
  )

# Count the number of entries below -1.2 m/s
below_threshold_1_2 <- df %>%
  filter(get(peak_braking_velocity_col) < -1.2) %>%
  summarise(Count_Below_Threshold = n())

# Filter dataset: Exclude values below -1.2 m/s
df_filtered <- df %>% filter(get(peak_braking_velocity_col) >= -1.2)

# Summary statistics AFTER exclusion
summary_after <- df_filtered %>%
  summarise(
    Remaining_Entries = n(),
    Mean_Braking_Velocity = mean(get(peak_braking_velocity_col), na.rm = TRUE),
    Median_Braking_Velocity = median(get(peak_braking_velocity_col), na.rm = TRUE),
    Min_Braking_Velocity = min(get(peak_braking_velocity_col), na.rm = TRUE),
    Max_Braking_Velocity = max(get(peak_braking_velocity_col), na.rm = TRUE),
    SD_Braking_Velocity = sd(get(peak_braking_velocity_col), na.rm = TRUE)
  )

# Print results
print("Summary before exclusion:")
print(summary_before)

print("Entries below -1.2 m/s:")
print(below_threshold_1_2)

print("Summary after exclusion:")
print(summary_after)

# OPTIONAL: Histogram to visualize distribution before/after exclusion
library(ggplot2)



```

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Create a data frame for plotting
comparison_df <- data.frame(
  Statistic = c("Mean Braking Velocity", "Median Braking Velocity", 
                "Min Braking Velocity", "Max Braking Velocity", 
                "SD Braking Velocity"),
  Before_Exclusion = c(-1.2268, -1.2247, -1.8996, -0.6871, 0.2266),
  After_Exclusion = c(-1.0336, -1.0498, -1.1998, -0.6871, 0.1237)
)

# Convert data to long format for ggplot
comparison_long <- tidyr::pivot_longer(comparison_df, 
                                       cols = c(Before_Exclusion, After_Exclusion),
                                       names_to = "Condition", 
                                       values_to = "Value")

# Create the bar plot
ggplot(comparison_long, aes(x = Statistic, y = Value, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Before_Exclusion" = "blue", "After_Exclusion" = "red")) +
  labs(title = "Comparison of Braking Velocity Statistics Before and After Exclusion (-1.2 m/s)",
       x = "Statistic",
       y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

```

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Define the peak braking velocity column
peak_braking_velocity_col <- "Peak Braking Velocity"

# Filter dataset (excluding values below -1.2 m/s)
df_filtered <- df %>% filter(get(peak_braking_velocity_col) >= -1.2)

# Calculate mean values before and after exclusion, handling NA values
comparison_df <- data.frame(
  Metric = c("Braking RFD", "Braking Net Impulse", "Peak Propulsive Power"),
  Before_Exclusion = c(mean(df$`Braking RFD`, na.rm = TRUE),
                        mean(df$`Braking Net Impulse`, na.rm = TRUE),
                        mean(df$`Peak Propulsive Power`, na.rm = TRUE)),
  After_Exclusion = c(mean(df_filtered$`Braking RFD`, na.rm = TRUE),
                       mean(df_filtered$`Braking Net Impulse`, na.rm = TRUE),
                       mean(df_filtered$`Peak Propulsive Power`, na.rm = TRUE))
)

# Drop any NA values that could cause missing bars
comparison_df <- comparison_df %>% drop_na()

# Convert data to long format for ggplot
comparison_long <- pivot_longer(comparison_df, 
                                cols = c(Before_Exclusion, After_Exclusion),
                                names_to = "Condition", 
                                values_to = "Value")

# Create the bar plot
ggplot(comparison_long, aes(x = Metric, y = Value, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Before_Exclusion" = "blue", "After_Exclusion" = "red")) +
  labs(title = "Comparison of Key Performance Metrics Before and After Exclusion (-1.2 m/s)",
       x = "Metric",
       y = "Mean Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

```

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Define the peak braking velocity column
peak_braking_velocity_col <- "Peak Braking Velocity"

# Filter dataset (excluding values below -1.2 m/s)
df_filtered <- df %>% filter(get(peak_braking_velocity_col) >= -1.2)

# Ensure column names match exactly
colnames(df)

# Handle NA values using `na.rm = TRUE`
comparison_df <- data.frame(
  Metric = c("L|R Peak Braking Force", "L|R Avg. Braking Force", 
             "L|R Braking Impulse Index", "L|R Peak Propulsive Force"),
  Before_Exclusion = c(mean(df$`L|R Peak Braking Force`, na.rm = TRUE),
                        mean(df$`L|R Avg. Braking Force`, na.rm = TRUE),
                        mean(df$`L|R Braking Impulse Index`, na.rm = TRUE),
                        mean(df$`L|R Peak Propulsive Force`, na.rm = TRUE)),
  After_Exclusion = c(mean(df_filtered$`L|R Peak Braking Force`, na.rm = TRUE),
                       mean(df_filtered$`L|R Avg. Braking Force`, na.rm = TRUE),
                       mean(df_filtered$`L|R Braking Impulse Index`, na.rm = TRUE),
                       mean(df_filtered$`L|R Peak Propulsive Force`, na.rm = TRUE))
)

# Convert data to long format for ggplot
comparison_long <- pivot_longer(comparison_df, 
                                cols = c(Before_Exclusion, After_Exclusion),
                                names_to = "Condition", 
                                values_to = "Value")

# Create the bar plot
ggplot(comparison_long, aes(x = Metric, y = Value, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Before_Exclusion" = "blue", "After_Exclusion" = "red")) +
  labs(title = "Comparison of Metrics with the Largest Changes Before and After Exclusion (-1.2 m/s)",
       x = "Metric",
       y = "Mean Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))


```

### **Interpretation of the Graph**  

- **L|R Peak Braking Force**   
  - Increased significantly after exclusion, showing that low-intent jumps had much weaker braking force.  
  - Athletes who showed intent applied far more force when resisting downward movement.  

- **L|R Avg. Braking Force** ️  
  - Became stronger after exclusion, meaning athletes in low-intent jumps weren’t applying consistent braking force.  
  - The remaining data represents a more **accurate** picture of engaged efforts.  

- **L|R Braking Impulse Index**  
  - Increased dramatically, meaning low-intent jumps had much **less force over time**.  
  - Athletes showing intent were absorbing and redirecting force more effectively.  

- **L|R Peak Propulsive Force**  
  - Increased slightly, meaning braking velocity impacts propulsive force but **not as much** as braking-related metrics.  
  - Some athletes still generated good takeoff power despite weak braking force.  

```{r}
# Load necessary libraries
library(dplyr)
library(broom)

# Define the peak braking velocity column
peak_braking_velocity_col <- "Peak Braking Velocity"

# Filter dataset (excluding values below -1.2 m/s)
df_filtered <- df %>% filter(get(peak_braking_velocity_col) >= -1.2)

# Add an indicator variable: Before = 0, After = 1
df_before <- df %>% 
  select(`L|R Peak Braking Force`, `L|R Avg. Braking Force`, 
         `L|R Braking Impulse Index`, `L|R Peak Propulsive Force`) %>%
  mutate(Exclusion = 0)

df_after <- df_filtered %>% 
  select(`L|R Peak Braking Force`, `L|R Avg. Braking Force`, 
         `L|R Braking Impulse Index`, `L|R Peak Propulsive Force`) %>%
  mutate(Exclusion = 1)

# Combine both datasets
df_combined <- bind_rows(df_before, df_after)

# Function to run regressions and extract results
run_regression <- function(dependent_var) {
  model <- lm(get(dependent_var) ~ Exclusion, data = df_combined)
  return(tidy(model))  # Get summary in a clean format
}

# Run regressions for each metric
metrics <- c("L|R Peak Braking Force", "L|R Avg. Braking Force", 
             "L|R Braking Impulse Index", "L|R Peak Propulsive Force")

# Store results in a list
regression_results <- lapply(metrics, run_regression)

# Convert list to a dataframe for easy viewing
regression_summary <- bind_rows(regression_results, .id = "Metric")

# Print results
print(regression_summary)

```

Significant Changes (p < 0.05):

L|R Avg. Braking Force (p = 0.004) → Exclusion significantly increased average braking force.
L|R Braking Impulse Index (p = 0.004) → Exclusion significantly increased braking impulse over time.

 Not Significant (p > 0.05):

L|R Peak Braking Force (p = 0.057) → Close, but not statistically significant.
L|R Peak Propulsive Force (p = 0.088) → No strong evidence of change after exclusion.


```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Define the new exclusion threshold (-1.0 m/s)
threshold_1_0 <- -1.0
peak_braking_velocity_col <- "Peak Braking Velocity"

# Filter dataset (excluding values below -1.0 m/s)
df_filtered <- df %>% filter(get(peak_braking_velocity_col) >= threshold_1_0)

# Count how many values are excluded vs retained
excluded_count <- df %>% filter(get(peak_braking_velocity_col) < threshold_1_0) %>% nrow()
included_count <- nrow(df_filtered)

# Create summary table
exclusion_summary <- data.frame(
  Metric = c("Total Entries", "Entries Below -1.0 m/s", "Remaining Entries"),
  Count = c(nrow(df), excluded_count, included_count)
)

# Print summary table
print(exclusion_summary)

# Define the three selected metrics
selected_metrics <- c("Braking RFD", "Braking Net Impulse", "Peak Propulsive Power")

# Calculate mean values before and after exclusion
comparison_df <- data.frame(
  Metric = selected_metrics,
  Before_Exclusion = c(mean(df$`Braking RFD`, na.rm = TRUE),
                        mean(df$`Braking Net Impulse`, na.rm = TRUE),
                        mean(df$`Peak Propulsive Power`, na.rm = TRUE)),
  After_Exclusion = c(mean(df_filtered$`Braking RFD`, na.rm = TRUE),
                       mean(df_filtered$`Braking Net Impulse`, na.rm = TRUE),
                       mean(df_filtered$`Peak Propulsive Power`, na.rm = TRUE))
)

# Convert to long format for ggplot
comparison_long <- pivot_longer(comparison_df, 
                                cols = c(Before_Exclusion, After_Exclusion),
                                names_to = "Condition", 
                                values_to = "Value")

# Create the first bar plot (Selected Metrics)
ggplot(comparison_long, aes(x = Metric, y = Value, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Before_Exclusion" = "blue", "After_Exclusion" = "red")) +
  labs(title = "Comparison of Key Performance Metrics Before and After Exclusion (-1.0 m/s)",
       x = "Metric",
       y = "Mean Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

# Identify the top 4 metrics that changed the most
before_means <- df %>% summarise(across(where(is.numeric), mean, na.rm = TRUE))
after_means <- df_filtered %>% summarise(across(where(is.numeric), mean, na.rm = TRUE))

# Calculate absolute percentage change
change_percentage <- abs((after_means - before_means) / before_means * 100)
change_percentage <- pivot_longer(change_percentage, cols = everything(), names_to = "Metric", values_to = "Percentage_Change")

# Get the top 4 changing metrics
top_4_metrics <- change_percentage %>% arrange(desc(Percentage_Change)) %>% slice(1:4) %>% pull(Metric)

# Calculate means for top 4 metrics
comparison_top4 <- data.frame(
  Metric = top_4_metrics,
  Before_Exclusion = sapply(top_4_metrics, function(m) mean(df[[m]], na.rm = TRUE)),
  After_Exclusion = sapply(top_4_metrics, function(m) mean(df_filtered[[m]], na.rm = TRUE))
)

# Convert to long format for ggplot
comparison_top4_long <- pivot_longer(comparison_top4, 
                                     cols = c(Before_Exclusion, After_Exclusion),
                                     names_to = "Condition", 
                                     values_to = "Value")

# Create the second bar plot (Top 4 Changing Metrics)
ggplot(comparison_top4_long, aes(x = Metric, y = Value, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("Before_Exclusion" = "blue", "After_Exclusion" = "red")) +
  labs(title = "Comparison of Top 4 Changing Metrics Before and After Exclusion (-1.0 m/s)",
       x = "Metric",
       y = "Mean Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

```

At -1.2 m/s exclusion: You removed 169 entries (everything < -1.2 m/s).
At -1.0 m/s exclusion: You removed 260 entries (everything < -1.0 m/s).

